---
title: "Statistics in Action - case study 1"
author: "Toxicity assessment of the MON810 maize"
output:
  html_document:
    fig_height: 3
    fig_width: 8
    number_sections: yes
    toc: no
  pdf_document:
    fig_height: 3
    fig_width: 8
    includes:
      before_body: macros.tex
    keep_tex: yes
    number_sections: yes
    toc: no
---
```{r}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(dplyr, ggplot2, tidyverse, pwr)
```

</br>


# Introduction

The dataset <ttt>dataMON810_2018.csv</ttt> consists of several measurements made during a subchronic toxicity study concerning the MON810 maize.

Biochemical parameters reflecting most physiological functions were measured two times (week 5 and 14), in particular through serum and urine chemistry, and hematology. Organ weights were measured at week 14.

The main objective of this study is to evaluate possible GMO effects on these parameters.

# Single comparison

   1. We consider the variable "CALCIUM".
First we load the data
```{r}
data.study <- read.csv('/Users/Daniel/Documents/Courses X/Statistics in Action/Homework 1/dataMON810_2018.csv') 
#data.study <- read.csv('https://github.com/daniel-wientjens/Stats-in-Action-Homework-1/blob/master/dataMON810_2018.csv')
```
   

a. test if the mean level of calcium for period 2 is the same for males and females.  </br>**Hint:** plot first the data and justify the test(s) to use.

```{r}
#Splitting the dataset into the two periods 
data.study.p1 <- data.study %>% filter(period == 1)
data.study.p2 <- data.study %>% filter(period == 2)

#Splitting the dataset of period 2 into male and female and plotting the result
data.study.p2.m <- data.study.p2 %>% filter(sex == 'M')
data.study.p2.f <- data.study.p2 %>% filter(sex == 'F')

new <- aggregate(data.study.p2$CALCIUM ~ data.study.p2$sex, FUN= "mean") 
colnames(new) <- c('sex', 'calcium')

ggplot()+
  geom_point(data = data.study.p2.m, aes(x=sex, y=CALCIUM, colour = 'blue'), show.legend = F) + 
  geom_point(data = data.study.p2.f, aes(x=sex, y=CALCIUM, colour = 'red'), show.legend = F) +
  geom_point(data = new, aes(x=sex, y=calcium), size=5, alpha=0.5)

boxplot(CALCIUM ~ sex, data = data.study, col = "blue")
```

```{r}
par(mfrow = c(1, 2))
hist(data.study.p2.f$CALCIUM, main='Histogram of calcium for females')
hist(data.study.p2.m$CALCIUM, main='Histogram of calcium for male')
```
```{r}
aggregate(data.study.p2$CALCIUM ~ data.study.p2$sex, FUN= "sd" )
aggregate(data.study.p2$CALCIUM ~ data.study.p2$sex, FUN= "mean" )
```
After all these plot we can see that the means and variances are probably both different between the two groups. However we should not rely on our eyes and do some tests! 
So first we check whether the variances differ between the male and female polulation by performing an F-test. 
Lastly we use that result to conduct a T-test since we need to know whether to leave the var.equal at it's default value or to change it to TRUE. 
```{r}
alpha<-0.05

x <- data.study[data.study$sex=="M","CALCIUM"]
y <- data.study[data.study$sex=="F","CALCIUM"]
var.test(x,y, alternative = 'two.sided')

t.test(x,y, conf.level=1-alpha, var.equal = FALSE)
#add the wilcox test since we do not assume anything with it, when we do the t-test we assume the distribution
```
As we can see the F-test has pointed out that the variances are different. This is then used in the 
Since the p-value is very small, we can reject the H0 and say that the means of the control males and females is different.

     b. Test for the males if the mean level of calcium is the same for period 1 and period 2.
Here we split the original dataset into the males for period 1 and period 2 and keep only the calcium column in order to do our t-test. Since the population is similar we set the variance of the two samples to be equal.
```{r}
x <- data.study %>% filter(sex == 'M', period == '1') %>% select('CALCIUM', 'sex')
y <- data.study %>% filter(sex == 'M', period == '2') %>% select('CALCIUM', 'sex')

aggregate(x$CALCIUM ~ x$sex, FUN= "sd" )
aggregate(y$CALCIUM ~ y$sex, FUN= "sd" )

var.test(x$CALCIUM,y$CALCIUM, alternative = 'two.sided')

t.test(x$CALCIUM,y$CALCIUM, conf.level = 1-alpha, var.equal = FALSE)
```

Since the p-value of the Welch Two Sample t-test is very small, we can reject the H0 and say that the means of the control males in period 1 and period 2 is different

     c. test for the males if the mean level of calcium for period 2 is the same for the control group and the MON810 group.
As per usual we split the dataset into the two groups we want to compare, so here we filter them to only contain males from period 2 and have the control group vs the MON810 group.
Next we perform a F-test to see whether the variances differ between the two groups and use the result for our final t-test.
```{r}
data.study.cont <- data.study %>% filter(sex == 'M', period == '2', regimen == 'control')
data.study.MON810 <- data.study %>% filter(sex == 'M',period == '2', regimen == 'MON810') 
var.test(data.study.cont$CALCIUM, data.study.MON810$CALCIUM,conf.level = 1-alpha)
t.test(data.study.cont$CALCIUM, data.study.MON810$CALCIUM, conf.level = 1-alpha, var.equal = FALSE)
```
Here we first see that the F-test tells us that the variances are different between the two groups due to the low p-value (0.037) with an alpha of 5%. 
As we can see the p-value of the Welch Two Sample t-test is very high (0.446), so we do not reject the H0 there. This translates into the fact that the means of the males in period 2 of the control and MON810 group do not differ significantly.

    d. What is the probability to detect a difference of 1 sd (one standard deviation) with only 10 animals per group? with 20 animal? How can we ensure to detect such difference with a probability of 80%?
First we have to calculate the standard deviation
```{r}
#pooled version
#sd(data.study.p2.m$CALCIUM %>% na.omit() %>% filter(regimen %in% c('MON810', 'control')))
bla <- data.study.p2.m[data.study.p2.m$regimen %in% c("MON810", "control"), ]
bla2 <- sd(bla$CALCIUM %>% na.omit)
bla2
#should look into pooled version a bit more Louis got 1.09 or something
```
So the two groups are comprised of 10 individuals each, so the number of degrees of freedom is 10+10-2=18

```{r}
df <-18
dt <- 1
1-pt(qt(1-alpha/2,df)-dt,df) + pt(qt(alpha/2,df)-dt,df) 
```
So the power of the test should increase by the sample size, but the dt stays the same since delta is the standard deviation/standard deviation=1


```{r}
df <-18
#dt <- 1
dt<- 1/sqrt((1/5))
#look at dt <- 1/(1/#nx + 1/#ny)
1-pt(qt(1-alpha/2,df)-dt,df) + pt(qt(alpha/2,df)-dt,df) 
#or
dt <- 1
pwr.t.test(n=20, d=dt, type="two.sample", alternative="two.sided", sig.level=alpha)[[4]]
```
Next we want to find the number of observations for which the detection of more than 1 sd is equal or higher than 80%.
For this we fill in the power function and set 
```{r}
pwr.t.test(power=0.8, d=dt, sig.level=alpha) 
```
So we have to set n=17




-----------------------------------------------------------------------------------------


```{r}
pt(q = -sd(bla$CALCIUM %>% na.omit), df =18)
```
So 0.2%
for the second part we only adjust the dataframe since the standard deviation stays the same.
degrees of freedom = 20+20-2 =38
```{r}
pt(q = -sd(bla$CALCIUM %>% na.omit), df =38)
```
so now we have a probability of this happening of 0.1%
Third part
We now have to calculate what the upper- and lowerbounds of the probability 
```{r}
pt((d.mean-bla2)/s.d.mean, dfw) + 1 - pt((d.mean+bla2)/s.d.mean, dfw)
```


     
```{r}
#std.er <- sqrt((std.er(data.study.cont$CALCIUM)/10)+(std.er(data.study.MON810$CALCIUM)/10))
```


-----------------------------------------------------------------------------------------






     
     
     e. Test for the males if the mean levels of calcium for period 2 of the control group and the MON810 are equivalent. The equivalence limits will be defined using the 6 reference groups as 
     *i)* one standard deviation of the 6 reference means. 
```{r}
data.study.cont <- data.study %>% filter(sex == 'M', period == '2', regimen == 'control')
data.study.MON810 <- data.study %>% filter(sex == 'M',period == '2', regimen == 'MON810') 
data.study.ref <- data.study %>% filter(sex == 'M',period == '2', regimen != 'control', regimen != 'MON810') %>%  select(regimen, CALCIUM) %>% na.omit

ref <- aggregate(data.study.ref$CALCIUM ~ data.study.ref$regimen, FUN = 'mean')

sd(ref$`data.study.ref$CALCIUM`)
```
```{r}
t.power = function(nsamp=c(10,10),nsim=1000,means=c(0,0),sds=c(1,1)){
    lower = qt(.025,df=sum(nsamp) - 2)
    upper = qt(.975,df=sum(nsamp) - 2)
    ts = replicate(nsim,
       t.test(rnorm(nsamp[1],mean=means[1],sd=sds[1]),
              rnorm(nsamp[2],mean=means[2],sd=sds[2]))$statistic)

    sum(ts < lower | ts > upper) / nsim
}

pt(q = -sd(calciumEMean$CALCIUM), df = dim(calciumE)[1]-6)
```
     
     *ii)* two standard deviations of the 6 reference means.
     
     f. Summarize and comment these results.
   
   2. Do the same analysis with the variable "DIRBILI" (direct bilirubin)

a)test if the mean level of calcium for period 2 is the same for males and females. 
Hint: plot first the data and justify the test(s) to use.

   Here we unfortunately cannot do the same test as before since this is categorical data
   Nevertheless we plot the data and see what we can learn from it.

```{r}
dirbili <- data.study %>% select(regimen,sex,period,DIRBILI) %>% na.omit()
dirbili$period <- as.factor(dirbili$period)
```

```{r}
#Splitting the dataset into the two periods 
dirbili.p1 <- dirbili %>% filter(period == 1)
dirbili.p2 <- dirbili %>% filter(period == 2)

#Splitting the dataset of period 2 into male and female and plotting the result
dirbili.p2.m <- dirbili.p2 %>% filter(sex == 'M')
dirbili.p2.f <- dirbili.p2 %>% filter(sex == 'F')

dirbili.p1.m <- dirbili.p1 %>% filter(sex == 'M')
dirbili.p1.f <- dirbili.p1 %>% filter(sex == 'F')

new <- aggregate(dirbili$DIRBILI ~ dirbili$sex, FUN= "mean") 
colnames(new) <- c('sex', 'dirbili')

ggplot() +
  geom_point(data = dirbili.p2.m, aes(x=sex, y=DIRBILI, colour = 'blue'), show.legend = F) + 
  geom_point(data = dirbili.p2.f, aes(x=sex, y=DIRBILI, colour = 'red'), show.legend = F) +
  geom_point(data = new, aes(x=sex, y=dirbili), size=5, alpha=0.5)
```

```{r}
table(dirbili.p2 %>% select(sex,DIRBILI))
```
Since this is categorical data and hence split into different classes we perform a chi squared test on the continuity table.
```{r}
chisq.test(table(dirbili.p2 %>% select(sex,DIRBILI)), correct = FALSE)
```
As we can see the p value is large and hence the H0 holds where we say that the means of the two groups male and female in period two do not differ.

b)test for the males if the mean level of dirbili is the same for period 1 and period 2.
```{r}
new <- aggregate(dirbili$DIRBILI ~ dirbili$period, FUN= "mean") 
colnames(new) <- c('period', 'dirbili')

ggplot() +
  geom_point(data = dirbili.p1.m, aes(x=period, y=DIRBILI, colour = period), show.legend = F) + 
  geom_point(data = dirbili.p2.m, aes(x=period, y=DIRBILI, colour = period), show.legend = F) +
  geom_point(data = new, aes(x=period, y=dirbili), size=5, alpha=0.5)
```


```{r}
table(dirbili %>% select(period,DIRBILI))
```
Since correction is only available for 2 by 2 tables we do not need to write it in the formula

```{r}
chisq.test(table(dirbili %>% select(period,DIRBILI)))
```
So we can reject the H0 hence we can say that the two means are different.


c)test for the males if the mean level of calcium for period 2 is the same for the control group and the MON810 group.
```{r}
dirbili.cont <- dirbili %>% filter(sex == 'M', period == '2', regimen == 'control')
dirbili.MON810 <- dirbili %>% filter(sex == 'M',period == '2', regimen == 'MON810') 
```

```{r}
new <- aggregate(dirbili.p2.m$DIRBILI ~ dirbili.p2.m$regimen, FUN= "mean")
colnames(new) <- c('regimen', 'dirbili')
new <- new %>% filter(regimen %in% c('control','MON810')) 

unique(dirbili.p2.m)

ggplot() +
  geom_point(data = dirbili.p2.m %>% filter(regimen == 'control'), aes(x=regimen, y=DIRBILI, colour = regimen), show.legend = F) + 
  geom_point(data = dirbili.p2.m %>% filter(regimen == 'MON810'), aes(x=regimen, y=DIRBILI, colour = regimen), show.legend = F) +
  geom_point(data = new, aes(x=regimen, y=dirbili), size=5, alpha=0.5)
```


</br>


# Multiple comparisons


   1. Redo the three tests of the previous section (questions a., b. and c.) for now comparing  the means of all the quantitative variables (see the annex 5 of the ANSES report <ttt>BIOT2009sa0285Ra.pdf</ttt> to know the type of each variable). Store the results (i.e. all the p-values) in a dataframe with one variable per row and four columns (name of the variable + three p-values). 
   
   2. For each of the three tests, adjust the p-values using the Bonferroni  and the Benjamini-Hochberg corrections. How can we interpret these results?
   


